\subsection{Implementation Details}
We implement FGTR with PyTorch and use CLIP ViT-B/32 as the image and text backbone. The input image is resized to $224\times224$, and the maximum text length is set to 77 tokens. Unless otherwise specified, we train the model with the bidirectional InfoNCE objective. For training augmentation, we apply random resized cropping, horizontal flipping, color jittering, Gaussian blurring, and random erasing to images. Text augmentation is enabled during training following the masked-token construction in the data loader.

For noisy correspondence evaluation, we follow the common synthetic noise protocol by randomly replacing a proportion of captions with mismatched captions. In the main Flickr30K setting, the noise ratio is set to 40\%, and the corrupted correspondence indices are loaded from the pre-generated noise file. The model is trained for 10 epochs with the first 5 epochs used as the warm-up stage. During warm-up, noisy samples predicted by the GMM are suppressed to avoid unstable early correction. After warm-up, the clean/noisy division is refreshed at the beginning of each epoch using the maintained per-sample loss bank.

We optimize the model with Adam using a learning rate of $1\times10^{-6}$, weight decay of $4\times10^{-5}$, and momentum parameters $(0.9,0.999)$. The batch size is set to 64, and the learning rate is scheduled by cosine decay. The temperature for contrastive learning is initialized to 0.02. We use PyTorch automatic mixed precision during training.

For FGTR-specific hyperparameters, the prototype memory size is set to 40,000, and prototype retrieval uses the top-5 nearest memory samples. The minimum memory size for enabling noisy-sample correction is 1,024. The GMM posterior threshold is set to 0.5 for both the initial full-dataset estimation and subsequent loss-bank based divisions. For visual context construction, the selected patch ratio is set to 0.4. Prototype confidence is computed from the entropy of the candidate prototype distribution, where the candidate logits are scaled by the learnable CLIP logit scale. We further combine this entropy-based confidence with the average image-image retrieval similarity. The sentence-token fusion coefficient is set to 1.0 in the reported setting.

All experiments are conducted on an NVIDIA RTX 4090 GPU server.
